% easychair.tex,v 3.5 2017/03/15

\documentclass{easychair}
%\documentclass[EPiC]{easychair}
%\documentclass[EPiCempty]{easychair}
%\documentclass[debug]{easychair}
%\documentclass[verbose]{easychair}
%\documentclass[notimes]{easychair}
%\documentclass[withtimes]{easychair}
%\documentclass[a4paper]{easychair}
%\documentclass[letterpaper]{easychair}

\usepackage{cleveref}
\usepackage{doc}

% use this if you have a long article and want to create an index
% \usepackage{makeidx}

\lstset{
  frame=none,
  xleftmargin=2pt,
  stepnumber=1,
  numbers=left,
  numbersep=5pt,
  numberstyle=\ttfamily\tiny\color[gray]{0.3},
  belowcaptionskip=\bigskipamount,
  escapeinside={*'}{'*},
  tabsize=2,
  emphstyle={\bf},
  commentstyle=\color{ForestGreen},
  stringstyle=\mdseries\rmfamily,
  showspaces=false,
  keywordstyle=\bfseries\rmfamily,
  columns=flexible,
  basicstyle=\small\sffamily,
  showstringspaces=false,
  morecomment=[l]\%,
}
\lstset{basicstyle=\footnotesize\ttfamily,breaklines=true}
% In order to save space or manage large tables or figures in a
% landcape-like text, you can use the rotating and pdflscape
% packages. Uncomment the desired from the below.
%
% \usepackage{rotating}
% \usepackage{pdflscape}

% Some of our commands for this guide.
%
\authorrunning{Abelt}

\newcommand{\easychair}{\textsf{easychair}}
\newcommand{\miktex}{MiK{\TeX}}
\newcommand{\texniccenter}{{\TeX}nicCenter}
\newcommand{\makefile}{\texttt{Makefile}}
\newcommand{\latexeditor}{LEd}


\newcommand{\LayeredTypes}{\textsc{LayeredTypes}}
%\makeindex

%% Front Matter
%%
% Regular title as in the article class.
%
\title{CAMELOT Research Scholarship -- Final Report}

% Authors are joined by \and. Their affiliations are given by \inst, which indexes
% into the list defined using \institute
%
\author{
Lukas Abelt\inst{1,2}
}

% Institutes for affiliations are also joined by \and,
\institute{
  LASIGE,
  Faculdade de Ciências da Universidade de Lisboa, Portugal\\
  \email{labelt@lasige.di.fc.ul.pt},
  \email{alcides@ciencias.ulisboa.pt}
\and
   Saarland University,
   Germany\\
}

%  \authorrunning{} has to be set for the shorter version of the authors' names;
% otherwise a warning will be rendered in the running heads. When processed by
% EasyChair, this command is mandatory: a document without \authorrunning
% will be rejected by EasyChair


% \titlerunning{} has to be set to either the main title or its shorter
% version for the running heads. When processed by
% EasyChair, this command is mandatory: a document without \titlerunning
% will be rejected by EasyChair
\titlerunning{Project Report CAMELOT}


\begin{document}
\maketitle



% The table of contents below is added for your convenience. Please do not use
% the table of contents if you are preparing your paper for publication in the
% EPiC Series or Kalpa Publications series

%\setcounter{tocdepth}{2}
%{\small
%\tableofcontents}

%\section{To mention}
%
%Processing in EasyChair - number of pages.
%
%Examples of how EasyChair processes papers. Caveats (replacement of EC
%class, errors).

%------------------------------------------------------------------------------
\section{Introduction}

This document is the final project report for the research scholarship granted in the scope of the CAMELOT project for the duration of 01.09.2022-30.11.2022. It was carried out by Lukas Abelt in the LASIGE Research Group within the Reliable Software Systems Research Line. The work was supercised by Alcides Fonseca, Assistant Professor at the Faculdade de Ciências da Universidade de Lisboa. The work was part of an Erasmus internship that ends in March 2023.

\section{Description of the Scholarship}
\subsection{Context}
\label{sec:context}
Most programmers and computer scientists will be familiar with simple type systems that ensure that the code they write is type safe in the context of the language they are writing it in. However, much more sophisticated type systems exist that can be used to ensure specific properties of a program, such as resource usage (Linear Types~\cite{linear}), value predicates (Liquid Types~\cite{DBLP:conf/pldi/RondonKJ08}), or correct following of a distributed protocol (Session Types~\cite{session}).

Some of these type systems can be dependent or independent from each other. Both Liquid Types and Session Types require a base type system with primitive types (such as int, bool) to exist, but they can be orthogonal to each other, as you can have a Liquid Type whose base type is int, and a Session Type that also relies on the base value being int, ignore the liquid predicate.

Given this plethora of type systems, different applications require a different subset of type systems. Not all applications require these advanced type systems (e.g., fully dependent types), trading off static guarantees for (even if short-term) productivity. Untyped languages have been popular in both web applications, scripting and data science. On the other hand, if you are writing a device driver, you will probably want to take advantage of a type system that guarantees that memory usage is constrained to a given bound~\cite{liquidate-assets}. Or if you are writing a complex, distributed protocol, you might want to take advantage of Session Types for that part of the program.

In this talk, we try to address the challenge of how can multiple type system co-exist in the same programming language, allowing different, valid combinations to be used in the type-checking of a program.
 
 
 To illustrate this in a small example, consider the program in \Cref{lst:code_before} that reads the text contents of a file and prints it to the console line-by-line. In this short snippet, there are multiple concerns to be addressed. Firstly, one wants to ensure that all function calls are done with variables of the appropriate data types. Secondly, to avoid errors due to an out of bound access, \texttt{get} should only be called with an index that does not exceed the list length. Lastly, when working with resources such as a file descriptor it might be useful to track it's state and ensure that it is in a proper (closed) state at the end of program execution. Note that the same standard library functions (\lstinline|createFD|, \lstinline|readLines|, etc...) may be needed in other programs with different type systems requirements.
 

These properties can be verified in two major ways. 

All features from all type systems can be combined in single, monolithic type system, containing all the complexity that all the different type systems entail. A very similar issue can also be observed when operating on Liquid Types: In certain cases we want to define predicates on orthogonal properties that could be verified independently from one another; However, current implementation of liquid types do not allow such distinction, leading to confusing and unintelligible error messages~\cite{wits-error-messages}. 



\begin{minipage}{0.4\linewidth}
\begin{lstlisting}[caption={Simple example code},label={lst:code_before}]
printLines lines idx len {
  if len != idx then {
    line = get(lines, idx)
    print(line)
    printLines(lines, idx+1, len)
  }
}

filename = "input.txt"
fileDescriptor = createFD(filename)

open(fileDescriptor)

lines = readLines(fileDescriptor)
len = length(lines)
printLines(lines, 0, len)

close(fileDescriptor)
\end{lstlisting}
\end{minipage}%
\begin{minipage}{0.59\linewidth}
\begin{lstlisting}[caption={Annotations for \LayeredTypes},label={lst:code_after}]
-- State Layer definitions
createFD :: state :: {} -> { Closed }
openFile :: state :: {Closed => Open} -> {}
readLines :: state :: {Open => Consumed} -> {}
closeFile :: state :: {Consumed => Closed} -> {}

-- Type layer definitions
get :: types :: List -> int -> string
length :: types :: List -> int
createFD :: types :: string -> FileHandle
open :: types :: FileHandle -> void
readLines :: types :: FileHandle -> List
close :: types :: FileHandle -> void
print :: types :: string -> void
printLines :: types :: List -> int -> int -> void

-- Liquid layer definitions
length :: {List | true} -> {v:int | v>=0}
printLines :: liquid :: { List | true } -> { l:int | l>=0 } -> { i:int | i<=l }
get :: liquid :: { List | true } -> { i:int | i<len }

-- State requirement at the end of the program
fileDescriptor :: state :: {Closed}	
\end{lstlisting}
\end{minipage}



\subsection{Goals}
\label{sec:proposed-approach}

The Goals of the Scholarship was to explore a new verification approach, called \LayeredTypes. The Goal was, that in \LayeredTypes, developers can write programs, but can also define additional type systems as layers. Each layer defines the basic types, and how typechecking happens. Additionally, a layer may depend on another layer if it requires information (such as types or typing contexts) from another layer.

Type checking of programs can be partial in the sense that a program may only use some layers (even if the libraries used are defined in more layers), requiring only that each function contains type information for that layer and all dependencies. The missing layers can be added over time, if they are found to be relevant.

For this purpose, a prototype was to be developed consisting of a small and simple functional language accompanied by a prototype layered Verification tool that can realise the described layered verification approach.

We believe that this layered, incremental approach can help build more powerful and independent type systems while at the same time making it simpler to understand the errors that might arise during verification.

\subsection{Results obtained}
During the duration of the scholarship a prototype compiler and verifier was built in a prototype language. This work layed an important foundation for the subsequent work for the remaining time of the ERASMUS+ Internship in which the implementation was further extended and improved. Two exemplary use-cases were developed showing the potential of the layered approach.

One example of how the framework works can be seen in the accompanging code fragments. In \Cref{lst:code_after} we see a set of annotations that can be added to the base code of \Cref{lst:code_before}. We define three separate layers \texttt{state}, \texttt{types} and \texttt{liquid}. Internally, the framework will build a dependency graph thus allowing to verify properties independently from one another where appropriate.

\subsection{Deliverables and Publications}

The full implementation of the developed prototype will be published on GitHub\footnote{\url{https://github.com/LuAbelt/LayeredTypes}}. In addition a submission for a contributed talk was made to the TYPES2023 conference\footnote{\url{https://types2023.webs.upv.es/Index.html}} that is still pending a notification of acceptance or rejection. Additionally it is planned to potentially submit a paper to a relevant conference later in 2023.

\input{../chapters/acknowledgements.tex}

\bibliographystyle{plain}
\bibliography{literature.bib}

\end{document}

%------------------------------------------------------------------------------
